---
title: "Step 4 — Extracting unique medication names"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(data.table)
library(tidyr)
library(stringr)
```

# Purpose

This file takes the `Prescription Items` field from the package-level dataset created in Step 3 and cleans it into individual medication names.

The medication text is standardised, split into individual drugs, and cleaned to remove dosage and formulation information.

Inputs:
- `packages_scripts_merged` from Step 3

Outputs:
- `prescription_drugs` (one row per `prescription_id` and cleaned drug name)
- `unique_drug_names_for_mapping` (distinct cleaned drug names used for the manual medication/disease-group mapping)


`unique_drug_names_for_mapping` is used for the manual mapping step, where each cleaned drug name is assigned a medication/disease group. The completed mapping is used in Step 5.


# Configuration

```{r config}
# Step 3 must be run first to create the package-prescription object.
if (!exists("packages_scripts_merged")) {
  stop("Run Step 3 first to create `packages_scripts_merged`.")
}

# Output file for the manual drug-mapping step.
drug_names_csv <- "unique_drug_names_for_mapping.csv"

```

```{r data table}
# package-level dataset created in Step 3.
full <- copy(packages_scripts_merged)
setDT(full)
```

# Normalise the Prescription Items text

One row per prescription, with the items string normalised: parenthetical notes removed, lower-cased, and all delimiters (commas, semicolons, slashes) unified to `;` 

```{r normalise-items}
meds <- full %>%
  select(prescription_id, prescription_items = `Prescription Items`) %>%
  mutate(
    prescription_items = str_remove_all(prescription_items, "\\s*\\(.*?\\)\\s*"), # drop "(...)" notes
    prescription_items = str_to_lower(prescription_items),
    prescription_items = str_replace_all(prescription_items, "[,;/]+", ";"),      # unify delimiters
    prescription_items = str_remove_all(prescription_items, "^;+|;+$"),           # trim stray ;
    prescription_items = na_if(prescription_items, "")
  ) %>%
  distinct(prescription_id, .keep_all = TRUE)
```

# Split into individual drugs and clean each name

`form_words` are dosage-form / packaging tokens (e.g. "tab", "inhaler", "suspension") that are not part of the drug name and are stripped out. Glued dose+form tokens (e.g. "5mgtab", "3mgtab") are removed as well.

```{r clean-drug-names}


form_words <- c(
  # tablet / capsule forms
  "tab", "tabs", "tablet", "tablets", "table",
  "cap", "capsule", "caps", "capsules",
  # other dosage forms
  "inh", "inhaler", "dose", "spray", "syrup", "ointment", "cream", "gel",
  "lotion", "suspension", "susp", "solution", "soln", "soluble", "liquid",
  "mixture", "injection", "inj", "vial", "ampoule", "neb", "nebule",
  "patch", "pessary", "suppository", "drop", "drops", "chew",
  # route / packaging descriptors
  "sl", "eye", "oral", "nasal", "wide mouth", "disp", "dispersable",
  # irregular fragments seen in the raw data 
  "mgtab", "mg tab", "2ug 0dose", "drops50u", "dtg -"
)
form_pattern <- paste0("(?<=^|\\s)(", paste(form_words, collapse = "|"), ")(?=\\s|$)")

meds_long <- meds %>%
  separate_rows(prescription_items, sep = ";|\\s+and\\s+") %>%
  mutate(
    med_raw = str_squish(prescription_items),
    drug_clean = med_raw %>%
      # dose + any glued form suffix, e.g. "500mg", "5ml", "5mgtab", "3mgtab", "5ml inj"
      str_remove_all("(?<![A-Za-z0-9])\\d+/?\\d*\\s*(?:mg|mcg|g|ml|%)\\s*(?:tabs?|tablets?|caps?|capsules?|inj|injection)?(?![a-z])") %>%
      str_remove_all(form_pattern) %>%                                              # form words
      str_remove_all("\\b(cr|xr)\\b") %>%                                           # CR/XR markers
      str_remove_all("(?<![A-Za-z])\\d+(?![A-Za-z])") %>%                           # stray numbers
      str_remove_all("\\.") %>%                                                     # stray dots
      str_squish()
  ) %>%
  # drop blanks and 1-2 character fragments (too short to be a real drug name).
  filter(drug_clean != "", nchar(drug_clean) > 2)
```

# Create mapping table

```{r save}
# one row per prescription_id and cleaned drug name.
prescription_drugs <- meds_long %>%
  distinct(prescription_id, drug_clean)

# distinct cleaned drug names for the manual mapping step.
unique_drug_names_for_mapping <- data.frame(
  drug_clean = sort(unique(prescription_drugs$drug_clean)),
  stringsAsFactors = FALSE
)

# export the unique drug list for manual classification.
write.csv(
  unique_drug_names_for_mapping,
  drug_names_csv,
  row.names = FALSE
)
```

# Quick look

```{r summary}

head(prescription_drugs)
head(unique_drug_names_for_mapping)

nrow(prescription_drugs)
nrow(unique_drug_names_for_mapping)
```
