---
title: "Step 5 — Assigning medication / disease groups"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# load in libraries
library(dplyr)
library(data.table)
library(tidyr)
library(stringr)
library(skimr)
library(readxl)
```

# Purpose

This file assigns the cleaned medication names produced in Step 4 to medication/disease groups using the manually completed `mapping_table.xlsx`.

Each cleaned drug name is matched to its assigned group, and the results are converted into binary indicators, with one column for each medication/disease group. A value of `1` indicates that the prescription contains at least one medication in that group, and `0` indicates that it does not.

Inputs:
- `prescription_drugs` from Step 4
- `packages_scripts_merged` from Step 3
- `mapping_table.xlsx` (manually completed mapping table)

Outputs:
- `medication_groups` (one row per package/prescription with binary medication/disease-group flags)
- `unmatched_medications` (cleaned drug names not found in the manual mapping table)

# Configuration

```{r config}
# Steps 3 and 4 must be run first to create the required R objects.
if (!exists("packages_scripts_merged")) {
  stop("Run Step 3 first to create `packages_scripts_merged`.")
}

if (!exists("prescription_drugs")) {
  stop("Run Step 4 first to create `prescription_drugs`.")
}

# Manually completed mapping table.
# This file is included in the repository.
mapping_xlsx <- "mapping_table.xlsx"
```

# Load inputs

```{r load}
# create working copies of the objects produced in earlier steps.
full <- copy(packages_scripts_merged)
items_clean <- copy(prescription_drugs)

# manually completed drug mapping.
mapping_raw <- read_excel(mapping_xlsx)
```

# Prepare the mapping table

```{r mapping}
# the 18 medication/disease groups expected in the final dataset.
all_groups <- c(
  "Cardiovascular conditions",
  "Central nervous system conditions",
  "Ear, nose and throat conditions",
  "Emergencies and injuries",
  "Endocrine conditions",
  "Eye conditions",
  "Family planning",
  "Gastro-intestinal conditions",
  "HIV and AIDS",
  "Infections and related conditions",
  "Kidney and urological disorders",
  "Mental health conditions",
  "Musculoskeletal conditions",
  "Nutrition and anaemia",
  "Obstetrics and Gynaecology",
  "Pain",
  "Respiratory conditions",
  "Skin conditions"
)

# clean the manually completed mapping into a two-column lookup table.
mapping <- mapping_raw %>%
  transmute(
    drug_clean = str_squish(str_to_lower(drug_clean)),
    group = str_squish(`Drug/Disease Group`)
  ) %>%
  filter(
    !is.na(drug_clean),
    drug_clean != "",
    !is.na(group),
    group != ""
  ) %>%
  distinct()

# check that every mapped group is one of the 18 expected groups.
unexpected_groups <- setdiff(
  unique(mapping$group),
  all_groups
)

if (length(unexpected_groups)) {
  warning(
    "Mapping contains groups not in all_groups: ",
    paste(unexpected_groups, collapse = "; ")
  )
}
```

# Identify unmatched medications (for review)

Drugs that appear in the prescription data but are not present in the mapping table do not receive a medication/disease group. These are retained in `unmatched_medications` for review.


```{r unmatched-report}
unmatched_medications <- items_clean %>%
  anti_join(mapping, by = "drug_clean") %>%
  count(drug_clean, name = "n_prescriptions", sort = TRUE)

message("Distinct unmatched drug names: ", nrow(unmatched_medications))

head(unmatched_medications, 30)
```

# Assign group flags

```{r flags}
# match each cleaned drug to its group, then create one binary column per group.
flags <- items_clean %>%
  inner_join(mapping, by = "drug_clean") %>%
  distinct(prescription_id, group) %>%
  mutate(
    value = 1L,
    group = factor(group, levels = all_groups)
  ) %>%
  pivot_wider(
    names_from = group,
    values_from = value,
    values_fill = 0L
  )
```

# Add group flags to package/prescription IDs

```{r assemble}
# retain one row for each package/prescription/patient combination.
base_ids <- full %>%
  select(id, prescription_id, patient_id
  ) %>% distinct()

medication_groups <- base_ids %>% left_join(flags, by = "prescription_id")

# any expected group that did not occur still needs a zero-filled column.
missing_cols <- setdiff(all_groups, names(medication_groups))

for (cc in missing_cols) {
  medication_groups[[cc]] <- 0L
}

# fix the final column order.
medication_groups <- medication_groups %>%
  select(
    id,
    prescription_id,
    patient_id,
    all_of(all_groups)
  )
```

# QC: prescriptions with no group at all

```{r qc-all-zero}
setDT(medication_groups)
n_all_zero <- medication_groups[, sum(rowSums(.SD != 0, na.rm = TRUE) == 0), .SDcols = all_groups]
message("Rows with no medication group assigned: ", n_all_zero)
```

# Quick look

```{r summary}

skim(medication_groups)

head(unmatched_medications)
```
