---
title: "Step 6 — Matching facilities & pickup points to coordinates"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(stringr)
library(tidyr)
library(readxl)
library(data.table)
library(openxlsx)
```

# Purpose

This file matches the distinct facilities and pickup points in the package-level dataset created in Step 3 to geographic coordinates.

Two coordinate reference sources are used:

- `PuPs_geolocation.xlsx`
- DHIS facility master list

The final matching approach uses exact-string matching in three passes. Each pass attempts to match names that remained unmatched after the previous one:

1. exact match on the combined reference name,
2. exact match on the DHIS short facility name,
3. exact match on the DHIS short facility name after removing the word "clinic" from both sides.

Each matched row records the matching `method` and coordinate `src` so that the process can be reviewed.

Rows that remain unmatched are exported to `locations_match_audit.xlsx`. Missing coordinates are then completed manually, producing the completed location mapping used in the next step.

Inputs:
- `packages_scripts_merged` from Step 3
- `PuPs_geolocation.xlsx` (raw reference file)
- DHIS facility master list (raw reference file)

Outputs:
- `fac_match_table` (R object containing facility coordinate matches)
- `pp_match_table` (R object containing pickup-point coordinate matches)
- `locations_match_audit.xlsx` (exported for manual review and completion)



# Configuration

```{r config}
# Step 3 must be run first to create the package-prescription object.
if (!exists("packages_scripts_merged")) {
  stop("Run Step 3 first to create `packages_scripts_merged`.")
}

# Paths to the raw coordinate reference files.
# The raw files are not included in this repository.
pup_geo_xlsx <- "......."
dhis_xlsx    <- "......."

# Sheet containing the DHIS facility master data.
dhis_sheet <- "DHIS Master data download (2)"

# Output for the manual coordinate-review step.
audit_out_xlsx <- "locations_match_audit.xlsx"
```

# Load inputs

```{r load}
# create a working copy of the package-level dataset from Step 3.
full <- copy(packages_scripts_merged)

# coordinate reference data.
locations  <- read_excel(pup_geo_xlsx)
facilities <- read_excel(dhis_xlsx, sheet = dhis_sheet)
```

# Build the combined reference table

Pickup-point and facility coordinates are stacked into one reference keyed on `ref_name`. Facility names have their leading two-letter province abbreviation stripped. Where the same name exists in both sources, the facility record is kept (DHIS is the more known source).

```{r reference-table}
# pickup-point geolocations.
locs_locations_long <- locations %>%
  transmute(
    ref_name  = Name,
    Longitude = suppressWarnings(as.numeric(Longitude)),
    Latitude  = suppressWarnings(as.numeric(Latitude)),
    src       = "locations"
  )

# DHIS facilities (strip the leading province abbreviation, e.g. "wc ").
locs_facilities_long <- facilities %>%
  transmute(
    ref_name_raw = `Facility Name`,
    Longitude    = suppressWarnings(as.numeric(longitude)),
    Latitude     = suppressWarnings(as.numeric(latitude)),
    src          = "facilities"
  ) %>%
  mutate(ref_name = str_remove(ref_name_raw, "^[a-z]{2}\\s+")) %>%
  select(ref_name, ref_name_raw, Longitude, Latitude, src)


# combine the pickup-point and DHIS facility coordinate references.
# keep only records with valid names and coordinates- if the same in both sources, keep the DHIS first
locs_ref_long <- bind_rows(locs_locations_long, locs_facilities_long) %>%
  filter(!is.na(Longitude), !is.na(Latitude), !is.na(ref_name), ref_name != "") %>%
  arrange(factor(src, levels = c("facilities", "locations"))) %>%  # facilities first
  distinct(ref_name, .keep_all = TRUE)                             # prefer facilities
```

# Helper functions

```{r helpers}

# remove a whole word from a name and clean up extra spaces -- clinic
# matching is case-insensitive
rm_word <- function(x, word = "clinic") {
  x %>%
    str_replace_all(paste0("(?i)\\b", word, "\\b"), "") %>%
    str_squish()
}


# return first value in a vector that is not missing and not blank
# used when several rows belong to the same site and we need to keep one usable admin label, such as province, district or sub-district
first_nm <- function(x) {
  ix <- which(!is.na(x) & x != "")
  if (length(ix)) x[ix[1]] else NA_character_
}

# match a set of site names to the coordinate reference table using exact name matching.
# this only matches names where Name is exactly the same as ref_name.
match_coords <- function(vec_names, ref_tbl) {
  target <- tibble(Name = vec_names) %>%
    distinct() %>%
    filter(!is.na(Name), Name != "")

  target %>%
    left_join(
      ref_tbl %>% mutate(ref_name_keep = ref_name, ref_name_raw_keep = ref_name_raw),
      by = c("Name" = "ref_name")
    ) %>%
    transmute(
      Name,
      matched_ref     = ref_name_keep,       # matched reference name
      matched_ref_raw = ref_name_raw_keep,   # original DHIS name (with province prefix)
      Longitude, Latitude, src,
      method = ifelse(!is.na(Longitude) & !is.na(Latitude), "raw_exact", NA_character_)
    ) %>%
    distinct(Name, .keep_all = TRUE)
}

# Add rescue matches to the base match table
# rescue matches are only used where the original exact match did not find coordinates
# existing successful matches are kept and are not overwritten
apply_rescue <- function(base_tbl, hits_tbl) {
  if (nrow(hits_tbl) == 0) return(base_tbl)
  base_tbl %>%
    left_join(
      hits_tbl %>% rename(
        matched_ref_hit     = matched_ref,
        matched_ref_raw_hit = matched_ref_raw,
        Longitude_hit       = Longitude,
        Latitude_hit        = Latitude,
        src_hit             = src,
        method_hit          = method
      ),
      by = "Name"
    ) %>%
    mutate(
      matched_ref     = coalesce(matched_ref,     matched_ref_hit),
      matched_ref_raw = coalesce(matched_ref_raw, matched_ref_raw_hit),
      Longitude       = coalesce(Longitude,       Longitude_hit),
      Latitude        = coalesce(Latitude,        Latitude_hit),
      src             = coalesce(src,             src_hit),
      method          = coalesce(method,          method_hit)
    ) %>%
    select(Name, matched_ref, matched_ref_raw, Longitude, Latitude, src, method)
}
```

# Distinct names to match (from the merged data)

```{r unique-names}
fac_unique <- full %>%
  filter(!is.na(Facility), Facility != "") %>%
  distinct(Facility) %>%
  rename(Name = Facility)

pp_unique <- full %>%
  filter(!is.na(`Pickup point`), `Pickup point` != "") %>%
  distinct(`Pickup point`) %>%
  rename(Name = `Pickup point`)
```

# Pass 1 — exact match on the combined reference name

```{r pass1}
fac_match_table <- match_coords(fac_unique$Name, locs_ref_long)
pp_match_table  <- match_coords(pp_unique$Name,  locs_ref_long)
```

# Short-name reference (used by passes 2 and 3)

The DHIS short facility name is used for the second and third matching passes.


```{r short-ref}
short_ref <- facilities %>%
  transmute(
    ref_name_full  = `Facility Name`,
    ref_name_short = `Short Facility Name`,
    Longitude      = suppressWarnings(as.numeric(longitude)),
    Latitude       = suppressWarnings(as.numeric(latitude))
  ) %>%
  filter(!is.na(ref_name_short), ref_name_short != "",
         !is.na(Longitude), !is.na(Latitude)) %>%
  distinct(ref_name_short, .keep_all = TRUE)
```

# Pass 2 — exact match on the DHIS short facility name

```{r pass2}
fac_unmatched <- fac_match_table %>% filter(is.na(Longitude) | is.na(Latitude)) %>% select(Name)
pup_unmatched <- pp_match_table  %>% filter(is.na(Longitude) | is.na(Latitude)) %>% select(Name)

fac_short_hits <- fac_unmatched %>%
  inner_join(short_ref, by = c("Name" = "ref_name_short")) %>%
  transmute(Name, Longitude, Latitude,
            matched_ref = Name, matched_ref_raw = ref_name_full,
            src = "facilities_short", method = "raw_exact_short")

pup_short_hits <- pup_unmatched %>%
  inner_join(short_ref, by = c("Name" = "ref_name_short")) %>%
  transmute(Name, Longitude, Latitude,
            matched_ref = Name, matched_ref_raw = ref_name_full,
            src = "facilities_short", method = "raw_exact_short")

fac_match_table <- apply_rescue(fac_match_table, fac_short_hits)
pp_match_table  <- apply_rescue(pp_match_table,  pup_short_hits)
```

# Pass 3 — short name with "clinic" removed from both sides

```{r pass3}
fac_unmatched <- fac_match_table %>% filter(is.na(Longitude) | is.na(Latitude)) %>% select(Name)
pup_unmatched <- pp_match_table  %>% filter(is.na(Longitude) | is.na(Latitude)) %>% select(Name)

fac_short_hits <- fac_unmatched %>%
  mutate(Name_nc = rm_word(Name, "clinic")) %>%
  inner_join(short_ref %>% mutate(ref_name_short_nc = rm_word(ref_name_short, "clinic")),
             by = c("Name_nc" = "ref_name_short_nc")) %>%
  transmute(Name, Longitude, Latitude,
            matched_ref = Name_nc, matched_ref_raw = ref_name_full,
            src = "facilities_short_nc", method = "raw_exact_short_nc") %>%
  distinct(Name, .keep_all = TRUE)

pup_short_hits <- pup_unmatched %>%
  mutate(Name_nc = rm_word(Name, "clinic")) %>%
  inner_join(short_ref %>% mutate(ref_name_short_nc = rm_word(ref_name_short, "clinic")),
             by = c("Name_nc" = "ref_name_short_nc")) %>%
  transmute(Name, Longitude, Latitude,
            matched_ref = Name_nc, matched_ref_raw = ref_name_full,
            src = "facilities_short_nc", method = "raw_exact_short_nc") %>%
  distinct(Name, .keep_all = TRUE)

fac_match_table <- apply_rescue(fac_match_table, fac_short_hits)
pp_match_table  <- apply_rescue(pp_match_table,  pup_short_hits)
```

# QC: unmatched names

Rows that remain unmatched are completed manually in the exported audit workbook.


```{r unmatched-count}
message("Facilities still unmatched: ",
        nrow(fac_match_table[is.na(fac_match_table$Longitude) | is.na(fac_match_table$Latitude), ]),
        " of ", nrow(fac_match_table))
message("Pickup points still unmatched: ",
        nrow(pp_match_table[is.na(pp_match_table$Longitude) | is.na(pp_match_table$Latitude), ]),
        " of ", nrow(pp_match_table))
```

# Attach administrative labels

Province, District, and Sub District are taken from the Step 3 data so that each site retains the administrative area under which it was recorded.


```{r admin-labels}
fac_admin_lookup <- full %>%
  filter(!is.na(Facility), Facility != "") %>%
  group_by(Facility) %>%
  summarise(Province_full    = first_nm(Province),
            District_full    = first_nm(District),
            Subdistrict_full = first_nm(`Sub District`),
            .groups = "drop")

pp_admin_lookup <- full %>%
  filter(!is.na(`Pickup point`), `Pickup point` != "") %>%
  group_by(`Pickup point`) %>%
  summarise(Province_full    = first_nm(Province),
            District_full    = first_nm(District),
            Subdistrict_full = first_nm(`Sub District`),
            .groups = "drop")

fac_match_table <- fac_match_table %>% left_join(fac_admin_lookup, by = c("Name" = "Facility"))
pp_match_table  <- pp_match_table  %>% left_join(pp_admin_lookup,  by = c("Name" = "Pickup point"))
```

# Export audit workbook

```{r save-audit}

# Rows with missing coordinates are completed manually.
# The completed workbook is used in the next pipeline step.
write.xlsx(
  list(
    Facilities = fac_match_table,
    PickupPoints = pp_match_table
  ),
  audit_out_xlsx,
  asTable = TRUE,
  overwrite = TRUE
)
```
