---
title: "Step 8 — Combining medication groups with the package data (dropping blank gender/supply)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# load in libraries
library(dplyr)
library(stringr)
library(tidyr)
library(readxl)
library(data.table)
library(skimr)
```

# Purpose


This file joins the medication-group flags created in Step 5 onto the package-level data with distances created in Step 7, then removes records that are too incomplete to use downstream.

Two exclusions are applied:

- rows with a blank `Gender`
- rows with a blank or `"NULL"` `Months Supplied` or `Prescription Length`

The supply fields are available for patients matched to the October extract during the Step 2 enrichment. This filter therefore restricts the dataset to records with complete gender and supply information. The number of rows removed by each exclusion is reported below.

Inputs:
- `packages_with_distance` from Step 7
- `medication_groups` from Step 5

Output:
- `package_level_clean` (cleaned package-level dataset used in the patient-level steps)


# Configuration

```{r config}
# Steps 5 and 7 must be run first to create the required R objects.
if (!exists("packages_with_distance")) {
  stop("Run Step 7 first to create `packages_with_distance`.")
}

if (!exists("medication_groups")) {
  stop("Run Step 5 first to create `medication_groups`.")
}
```

# Load inputs

```{r load}
# create working copies of the objects produced in earlier steps.
full_with_distance <- copy(packages_with_distance)
med_groups <- copy(medication_groups)

setDT(full_with_distance)
setDT(med_groups)
```

# Apply the exclusions

```{r drop-blank-gender}
# Drop rows with no Gender.
n_before <- nrow(full_with_distance)
full_with_distance <- full_with_distance[!is.na(Gender) & trimws(as.character(Gender)) != ""]
message("Dropped blank-gender rows: ", n_before - nrow(full_with_distance))
```

```{r drop-blank-supply}
# Treat NA, "", and "NULL" (any case) as blank - the original code relied on
# data.table dropping NA rows; this makes that behaviour explicit.
is_blank <- function(x) {
  x <- trimws(as.character(x))
  is.na(x) | x == "" | toupper(x) == "NULL"
}

n_before <- nrow(full_with_distance)
full_with_distance <- full_with_distance[
  !is_blank(`Months Supplied`) & !is_blank(`Prescription Length`)
]
message("Dropped blank Months Supplied / Prescription Length rows: ",
        n_before - nrow(full_with_distance))
```

```{r qc-values}
# Sanity check the fields we filtered on.
unique(full_with_distance$Gender)
sort(unique(full_with_distance$`Months Supplied`))
sort(unique(full_with_distance$`Prescription Length`))
```

# Join the medication-group flags

```{r join-med-groups}
# med_groups holds id, prescription_id, patient_id, and the 18 group flag columns.
# Take only the group-flag columns that are not already in the package data.
group_cols <- setdiff(
  names(med_groups),
  c("id", names(full_with_distance))
)

# One row per package id.
med_groups <- unique(med_groups, by = "id")

# Align medication-group values to full_with_distance by package id.
group_vals <- med_groups[
  full_with_distance,
  on = "id",
  ..group_cols
]

package_level_clean <- copy(full_with_distance)

for (nm in names(group_vals)) {
  package_level_clean[, (nm) := group_vals[[nm]]]
}

# prescriptions with no mapped medication group should have 0, not NA.
for (nm in group_cols) {
  set(package_level_clean, which(is.na(package_level_clean[[nm]])), nm, 0L)
}
```

# Drop leftover helper columns

```{r drop-helpers}
# Engineering/QC columns that leaked in from earlier steps (e.g. `is_dup` from
# Step 3). Dropped defensively: only those actually present are removed, so this
# never errors if an upstream step already cleaned them up.
helper_cols <- intersect(
  c("is_dup", "year_band", "pref_nonempty", "items_clean", "items_len"),
  names(package_level_clean)
)

if (length(helper_cols)) {
  package_level_clean[, (helper_cols) := NULL]
}
```

# QC: medication-group coverage

```{r qc-med-coverage}
# The 18 expected group columns
all_groups <- c(
  "Cardiovascular conditions", "Central nervous system conditions",
  "Ear, nose and throat conditions", "Emergencies and injuries",
  "Endocrine conditions", "Eye conditions", "Family planning",
  "Gastro-intestinal conditions", "HIV and AIDS",
  "Infections and related conditions", "Kidney and urological disorders",
  "Mental health conditions", "Musculoskeletal conditions",
  "Nutrition and anaemia", "Obstetrics and Gynaecology", "Pain",
  "Respiratory conditions", "Skin conditions"
)

# all 18 group columns must be present so a renamed or missing group column
# cannot pass through silently.
missing_group_cols <- setdiff(all_groups, names(package_level_clean))
stopifnot(length(missing_group_cols) == 0)

# rows with no medication group (all 18 flags 0): the prescription's drugs did not map, or its Prescription Items was empty. We KEEP these - an all-zero vector is a valid "none of these groups" state - and just report how many.
n_total    <- nrow(package_level_clean)
n_all_zero <- package_level_clean[, sum(rowSums(.SD != 0, na.rm = TRUE) == 0), .SDcols = all_groups]
message("Rows with no medication group (all 18 flags 0): ", n_all_zero,
        " of ", n_total, " (", round(100 * n_all_zero / n_total, 2), "%) - kept")
```


# Quick look

```{r summary}

skim(package_level_clean)
```
