---
title: "Step 1 — Formatting the October 2025 prescription extract"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# load in libraries
library(data.table)
library(tidyr)
library(future.apply)
```

# Purpose

This file takes the prescription level data received in October 2025 and reshapes it into a dataframe with one row per prescription.

A custom parser is used and does the following:

1. read every field as plain text
2. then uses the script date as an anchor to
   tell the address block (which comes before the date) apart from the
   structured fields (which come after the date).
   

Input - October 2025 prescription extract (raw)
Output - R object, `synch_all` (one row per October prescription)


# Configuration

```{r config}

# Path to the raw October 2025 extract.
# The raw data are not included in this repository.
zip_path <- "......."

# Folder containing the prescription CSVs within the source zip.
inner_dir <- "Synchsd-4206 Nikhil Oct 2025"  # subject to change

```

# The parser

```{r parser}

# Final output columns:
#   Province, District, Sub District, Registered date, Facility, Pickup point,
#   Patient address, Prescription Date, Prescription Status, patient_id,
#   prescription_id, Gender, date of birth, Months Supplied,
#   Prescription Length, Prescription Items
#
# Layout of a the raw input
#   1st 6 cols are fixed with comma separator 
#   7th is the address (varying number of commas that form part of the address) 
#   8th Prescription Date
#   9 Structured fields
#   Last: Prescription Items 


# function input: p is the file path to one raw prescription CSV

normalize_synch_fast <- function(p) {
  
  dt <- data.table::fread(
    p,                      
    sep        = ",",                  # set comma as the separator to start with 
    header     = TRUE,
    quote      = "",                   # do not treat quotation marks as start and end of a field 
    fill       = Inf,                  # avoid early stopping on too many fields
    colClasses = "character",          # read everything as character 
    na.strings = c("", "NA", "N/A"),   # treat all of these as missing values
    showProgress = FALSE               
  )
  if (!nrow(dt)) return(data.table())  # adding to stop if the current file happens to be blank 

  
  # make sure the table has at least 6 columns before renaming them.
  # if the file has fewer than 6 columns overall, add empty placeholder columns.
  if (ncol(dt) < 6L) {          
    for (k in (ncol(dt) + 1L):6L) dt[[k]] <- NA_character_  
  }
  # rename the first 6 columns by position, since these should always be the fixed location/date fields.
  setnames(dt, old = names(dt)[1:6],
           new = c("Province", "District", "Sub District",
                   "Registered date", "Facility", "Pickup point")) 
  

  #### everything from column 7 onward is the section we have to reshape ################################################################
  
  # count the number of cols separated by commas so that we can go from 7 onwards 
  nc <- ncol(dt)
  
  
  # if only the 6 cols from above exists, return the fixed fields and add the remaining fields as blank
  if (nc < 7L) {
    

    out <- dt[, .(
      Province, District, `Sub District`, `Registered date`, Facility, `Pickup point`,
      `Patient address`     = NA_character_,
      `Prescription Date`   = NA_character_,
      `Prescription Status` = NA_character_,
      patient_id            = NA_character_,
      prescription_id       = NA_character_,
      Gender                = NA_character_,
      `date of birth`       = NA_character_,
      `Months Supplied`     = NA_character_,
      `Prescription Length` = NA_character_,
      `Prescription Items`  = NA_character_
    )]
    out[, source_file := basename(p)]
    return(out[])
  }

  
  # find the tail columns
  tail_cols <- 7L:nc
  tail_mat  <- as.matrix(dt[, ..tail_cols])
  n         <- nrow(dt)
  
  
  # tail_mat now contains the imported columns from col 7 onward
  # this is converted to a matrix so we can search across each row by position 

  # create a pattern to identify dates in the tail columns
  # this matches observed dates in YYYY-MM-DD format, optionally followed by a time
  # the year is restricted to 2021-2025 based on the know data range.
  date_re   <- "^202[1-5]-\\d{2}-\\d{2}(?:\\s+\\d{2}:\\d{2}:\\d{2}(?:\\.\\d+)?)?$"
  
  # check each value in the imported tail section and mark whether it looks like a date.
  date_mask <- apply(tail_mat, 2L, function(x) grepl(date_re, x))
  
  # date_mask should be a matrix so row/column indexing works
  # if apply() simplifies the result, replace it with a same-sized FALSE matrix.
  if (!is.matrix(date_mask)) {
    date_mask <- matrix(FALSE, nrow = nrow(tail_mat), ncol = ncol(tail_mat))
  }


  # find the position of the first date in each tail row.
  # the address ends right before this date.
  # max.col() always returns a position, even if no TRUE/date is found.
  # to handle that, add a fake FALSE column at the start and subtract 1 to make sure 
  # the rows with no dates are at position 0
  pos <- max.col(cbind(FALSE, date_mask), ties.method = "first") - 1L

  
  
  # patient address = the tail before pos
  addr <- ifelse(
    pos == 0L,
    # there is no date and we treat the whole tail as the address (unlikely)
    apply(tail_mat, 1L, function(z) {
      z <- z[!is.na(z) & nzchar(trimws(z))]
      if (length(z)) paste(z, collapse = ",") else NA_character_
    }),
    
    # if pos is greater than 0, a date was found in the tail row.
    # the address is made from the values before that date.
    vapply(seq_len(n), function(i) {
      # if the date is the first value in the tail, there is no address before it
      if (pos[i] <= 1L) return(NA_character_)
      
       # take all tail values before the date position
      z <- as.character(tail_mat[i, 1:(pos[i] - 1L)])
      
       # remove missing and blank address pieces
      z <- z[!is.na(z) & nzchar(trimws(z))]
      
      # join the remaining address pieces back together with commas
      # if there is nothing left, return NA
      if (length(z)) paste(z, collapse = ",") else NA_character_
    }, FUN.VALUE = character(1))
  )

  # pull out the prescription date itself
  # if no date was found in the row, set Prescription Date to NA
  # pmax(pos, 1L) is only there to keep the matrix index valid, because matrix positions cannot be 0
  pdate <- ifelse(pos == 0L, NA_character_,
                  tail_mat[cbind(seq_len(n), pmax(pos, 1L))])

  # collect everything after the prescription date for each row
  # this should contain the 7 structured fields first, followed by Prescription Items
  after_list <- lapply(seq_len(n), function(i) {
    
    # if no date was found, there is no reliable prescription section to extract
    if (pos[i] == 0L) return(character(0))
    
    # start one position after the date
    j1 <- pos[i] + 1L
    
    # if the date is the last value in the tail, there is nothing after it
    if (j1 > ncol(tail_mat)) character(0)
    else as.character(tail_mat[i, j1:ncol(tail_mat)])
  })

  # pull the kth value after the prescription date from each row
  # this is used for the structured fields that always appear in the same order.
  take_field <- function(k) {
    vapply(after_list, function(v) {
      
      
      # if the row has at least k values after the date, and the kth value is not blank,
      # return that value with extra spaces removed
      if (length(v) >= k && nzchar(trimws(v[k]))) trimws(v[k]) else NA_character_
    }, FUN.VALUE = character(1))
  }
  
  # the first 7 values after the prescription date are the structured fields
  presc_status    <- take_field(1)
  patient_id      <- take_field(2)
  prescription_id <- take_field(3)
  gender          <- take_field(4)
  dob             <- take_field(5)
  months_sup      <- take_field(6)
  presc_len       <- take_field(7)

  # anything after the 7 structured fields is Prescription Items
  # join these pieces back together because item text may also contain commas
  items <- vapply(after_list, function(v) {
    if (length(v) > 7L) {
      
       # take values 8 onward, remove extra spaces, and drop blanks
      z <- trimws(v[8L:length(v)])
      z <- z[nzchar(z)]
      
      # join remaining item pieces back together
      if (length(z)) paste(z, collapse = ", ") else NA_character_
    } else NA_character_
  }, FUN.VALUE = character(1))

  # build the final cleaned table with one row per prescription
  out <- data.table(
    Province = dt[[1]], District = dt[[2]], `Sub District` = dt[[3]],
    `Registered date` = dt[[4]], Facility = dt[[5]], `Pickup point` = dt[[6]],
    `Patient address`     = trimws(addr),
    `Prescription Date`   = pdate,
    `Prescription Status` = presc_status,
    patient_id            = patient_id,
    prescription_id       = prescription_id,
    Gender                = gender,
    `date of birth`       = dob,
    `Months Supplied`     = months_sup,
    `Prescription Length` = presc_len,
    `Prescription Items`  = items
  )
  out[, source_file := basename(p)]
  out[]
}
```

# Extract and parse all October CSVs

```{r extract-and-parse}

# list the csvs inside the target folder of the zip.
zip_idx     <- utils::unzip(zip_path, list = TRUE)
csv_entries <- zip_idx$Name[grepl(paste0("^", inner_dir, "/.*\\.csv$"), zip_idx$Name)]
stopifnot(length(csv_entries) > 0)

# extract just those csvs to a temp folder.
exdir <- tempfile("synchsd_")
dir.create(exdir, recursive = TRUE, showWarnings = FALSE)
utils::unzip(zip_path, files = csv_entries, exdir = exdir)
csv_paths <- file.path(exdir, csv_entries)

# parse every CSV in parallel and stack the results.
plan(multisession, workers = max(1L, parallel::detectCores() - 1))
dt_list   <- future_lapply(csv_paths, normalize_synch_fast, future.seed = TRUE)
plan(sequential)
synch_all <- data.table::rbindlist(dt_list, use.names = TRUE, fill = TRUE)

# Finalise the R object
synch_all[, source_file := NULL]
```


# Quick look

```{r summary}
skimr::skim(synch_all)
```

