---
title: "Step 2 — May prescription set: de-duplicate, then enrich with October demographics"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load in libraries
library(data.table)
library(tidyr)
library(skimr)
library(tidyverse)
library(lubridate)
library(archive)
library(scales)
```

# Purpose

- The October 2025 prescription extract contains one script per patient, with complete patient-level details.

- The May 2025 extract has fewer columns but includes multiple scripts per patient, covering financial years 2021–22, 2022–23, and 2023–24.

- This file build the base prescription set from the May extract

- It includes all prescription records (ids, dates, facility, pickup point).
Resolve duplicate prescription_ids so each appears only once, including duplicates found across different CSV files.

- Add patient-level fields from the October 2025 extract by joining on patient_id.

- Add the four fields missing from May: Gender, Date of Birth, Months Supplied, and Prescription Length.
Add Prescription Items from October, so all patient-level descriptors are consistent and come from the October snapshot.

- This is not a row-wise merge; May provides prescription records, October provides patient-level details.

- Prescription Items, prescription length, and months supplied are all taken from the October extract for consistency.

- Medication information reflects each patient’s regimen in October 2025, used as a proxy for earlier prescriptions since CCMDD regimens are generally stable.

Inputs : `synch_all` (from Step 1) and May prescription extract (raw)
         
Outputs:
- `may_scripts_dedup` (one row per May prescription)
- `scripts_may_october_combined` (May prescriptions enriched with October patient-level fields)

# Configuration

```{r config}
# Path to the raw May 2025 extract.
# The raw data are not included in this repository.
may_zip <- "......."

# Step 1 must be run first to create the October prescription object.
if (!exists("synch_all")) {
  stop("Run Step 1 first to create `synch_all`.")
}

```

# 1. Read the May extract

Only keep the three financial years of interest; the package data we link to later runs from roughly mid-2021 to mid-2023.

```{r extract-may}
zip_files <- archive(may_zip)$path
want      <- grep("2021-22|2022-23|2023-24", zip_files, value = TRUE)

tmpdir <- file.path(tempdir(), "synch_extract")
dir.create(tmpdir, showWarnings = FALSE)
archive_extract(may_zip, files = want, dir = tmpdir)
```

```{r read-may}
# these csvs have a free-text final column (`Prescription Items`) full of unquoted commas. 
# the first 10 fields are well-formed; split on exactly the first 10 commas and treat everything after that as the Prescription Items 
read_one <- function(fname, path) {
  fpath      <- file.path(path, fname)
  lines      <- readLines(fpath, encoding = "UTF-8")

  header     <- strsplit(lines[1], ",", fixed = TRUE)[[1]]
  data_lines <- lines[-1]

  # first 10 comma-separated fields.
  first10 <- tstrsplit(data_lines, ",", fixed = TRUE, keep = 1:10)
  dt      <- as.data.table(first10)
  setnames(dt, header[1:10])

  # everything after the 10th comma -> Prescription Items (commas kept).
  med_col <- header[11]  # "Prescription Items"
  dt[, (med_col) := sub("^([^,]*,){10}(.*)$", "\\2", data_lines)]

  dt[, source_file := fname]
  dt
}

# read all selected May files and combine them into one table.
lst <- lapply(want, read_one, path = tmpdir)
all_may_prescriptions <- rbindlist(lst, fill = TRUE)
```

```{r fix-dates}
# normalise the two date columns to IDate (drop any trailing time component).
all_may_prescriptions[, `Registered date`   := as.IDate(sub(" .*", "", `Registered date`))]
all_may_prescriptions[, `Prescription Date`  := as.IDate(`Prescription Date`)]
```

```{r summary-may}
skim(all_may_prescriptions)
```

# 2. Investigate duplicate prescription_ids (QC)

The same `prescription_id` can appear in more than one financial-year file. Need to check  which columns actually differ across those duplicates before deciding how to resolve them.

```{r qc-which-cols-differ}

# find all rows where the same prescription_id appears more than once.
# keeps both the first and later copies of each duplicated prescription_id.
dup <- all_may_prescriptions[
  duplicated(prescription_id) | duplicated(prescription_id, fromLast = TRUE)
]


# for duplicated prescription_ids, check how many unique values each column has.
cols <- setdiff(names(dup), "prescription_id")

# convert the result to long format so each row isprescription_id, column name, number of unique values in that column.
tmp  <- dup[, lapply(.SD, uniqueN), by = prescription_id, .SDcols = cols]
tmp  <- melt(tmp, id.vars = "prescription_id",
             variable.name = "column", value.name = "n_unique")


# keep only columns where duplicated prescription_ids have conflicting values & count how many prescription_ids disagree for each column.
disagree_summary <- tmp[n_unique > 1, .N, by = column][order(-N)]
disagree_summary
```

`source_file` differs for essentially all duplicates, which is expected: the same prescription is pulled into more than one financial-year archive. The remaining duplicates (~16.8k) are genuine repeats we need to collapse.

```{r qc-switch-dates}
# sanity check: prescriptions whose dispense/prescription date precedes registration. Inspect only - not removed here.
switch_dates <- all_may_prescriptions[`Prescription Date` < `Registered date`]
nrow(switch_dates)
```

```{r qc-dup-counts}
dup_prescript_counts <- all_may_prescriptions[, .N, by = prescription_id][N > 1]
message("Unique prescription_ids with >1 row: ", nrow(dup_prescript_counts))
dup_prescript_counts
```

```{r qc-col-diff-detail}
# for the duplicated IDs, count how many IDs show real variation in each column (treating "" as NA so blanks don't count as a difference).
dup_ids <- dup_prescript_counts$prescription_id
diag_cols <- setdiff(names(all_may_prescriptions), c("prescription_id", "source_file"))

diff_flags <- all_may_prescriptions[
  prescription_id %in% dup_ids,
  lapply(.SD, function(x) uniqueN(na.omit(ifelse(x == "", NA, x))) > 1),
  by = prescription_id, .SDcols = diag_cols
]
col_diff_counts <- diff_flags[, lapply(.SD, sum), .SDcols = diag_cols]
col_diff_long   <- melt(col_diff_counts, measure.vars = diag_cols,
                        variable.name = "column",
                        value.name = "n_presc_ids_with_diff")[order(-n_presc_ids_with_diff)]
col_diff_long
```

# 3. Resolve duplicates to one row per prescription

For each `prescription_id` we keep a single "winner", chosen by priority:

1. most recent financial year (`year_band`: 2023-24 > 2022-23 > 2021-22),
2. a non-empty `Prescription Items` over an empty one, then
3. the longest `Prescription Items` string (most complete medication list).

(The `Prescription Items` tiebreakers only pick a representative, most-complete May row deterministically. The items themselves are replaced by October's in Step 4, so here they affect only which row's dates / facility / pickup point are kept.

```{r resolve-duplicates}
data <- copy(all_may_prescriptions)
setDT(data)

# priority 1: financial year (higher = more recent).
data[, year_band := fifelse(grepl("2023-24", source_file, ignore.case = TRUE), 3L,
                    fifelse(grepl("2022-23", source_file, ignore.case = TRUE), 2L,
                    fifelse(grepl("2021-22", source_file, ignore.case = TRUE), 1L, 0L)))]

# priority keys 2 & 3: prefer a populated, longer medication list.
data[, items_clean   := gsub("\\s+", " ", trimws(`Prescription Items`))]
data[, pref_nonempty := nzchar(items_clean)]
data[, items_len     := nchar(items_clean)]

# sort so the winner is first within each prescription_id, then take it.
setorder(data, prescription_id, -year_band, -pref_nonempty, -items_len)
may_scripts_dedup  <- data[, .SD[1], by = prescription_id]

# drop the helper columns so they don't leak downstream.
may_scripts_dedup [, c("year_band", "items_clean", "pref_nonempty", "items_len") := NULL]

# confirm that each prescription_id appears only once
stopifnot(sum(duplicated(may_scripts_dedup$prescription_id)) == 0)

```

# 4. Enrich the May set with October patient-level fields

```{r standardise-patient-id}

# create working copies for the enrichment step.
may_enriched <- copy(may_scripts_dedup)
october_data <- copy(synch_all)

# patient_id is the join key; lower-case and trim both sides so
# casing or spacing differences do not break the match.
may_enriched[, patient_id := str_trim(str_to_lower(patient_id))]
october_data[, patient_id := str_trim(str_to_lower(patient_id))]

# how many May patients are present in the October extract?
length(intersect(may_enriched$patient_id, october_data$patient_id))

```

```{r enrich-from-october}

# we take Prescription Items from October, so drop the May copy first.
may_enriched[, `Prescription Items` := NULL]

# patient-level fields supplied by October.
target_cols <- c(
  "date of birth",
  "Months Supplied",
  "Prescription Length",
  "Gender",
  "Prescription Items"
)

# the October extract should contain one row per patient_id.
stopifnot(sum(duplicated(october_data$patient_id)) == 0)

nsd <- october_data[
  , c("patient_id", intersect(target_cols, names(october_data))),
  with = FALSE
]

setkey(nsd, patient_id)
setkey(may_enriched, patient_id)

# left join: keep every May prescription and add October fields where available.
scripts_may_october_combined <- merge(
  may_enriched,
  nsd,
  by = "patient_id",
  all.x = TRUE
)

# row count must be unchanged by enrichment.
stopifnot(
  nrow(scripts_may_october_combined) == nrow(may_enriched)
)
```


```{r summary-final}
skim(scripts_may_october_combined)
```