---
title: "Step 3 — Linking prescriptions to dispensed packages"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(vroom)
library(dplyr)
library(data.table)
library(knitr)
library(tidyr)
library(skimr)
library(tidyverse)
library(lubridate)
library(archive)
library(scales)
library(stringr)
```

# Purpose

This file attaches the prescription information built in Step 2 to the package-level CCMDD data, which contains one row per package.

The join is a left join on `prescription_id`, keeping every package row and adding the matching prescription columns. Because one prescription can generate several packages, this is a one-to-many link from prescriptions to packages.

Inputs:
- `scripts_may_october_combined` from Step 2
- CCMDD package extract (raw)

Output:
- `packages_scripts_merged` (one row per package with prescription-level information added)

# Configuration

```{r config}

# Path to the raw CCMDD package extract.
# The raw data are not included in this repository.
ccmdd_csv <- "......."

# Step 2 must be run first to create the prescription object.
if (!exists("scripts_may_october_combined")) {
  stop("Run Step 2 first to create `scripts_may_october_combined`.")
}

```


```{r load}
# package-level data from the raw extract.
ccmdd <- fread(ccmdd_csv)

# prescription set created in Step 2.
prescription_data <- copy(scripts_may_october_combined)
```

# Duplicate checks (QC)

```{r qc-dups}

# double check for identical package rows.
sum(duplicated(ccmdd))

# packages that are identical once id/barcode are ignored
# (i.e. content duplicates differing only by their unique keys).
ccmdd <- as.data.table(ccmdd)

ccmdd[
  , is_dup := duplicated(.SD),
  .SDcols = setdiff(names(ccmdd), c("id", "barcode"))
]

ccmdd[, sum(is_dup)]
dup_dt <- ccmdd[is_dup == TRUE]

# remove the temporary duplicate flag before continuing.
ccmdd[, is_dup := NULL]

# fully identical prescription rows.
sum(duplicated(prescription_data))

```
We see no concerning dupliate issues


# Join packages to prescriptions

```{r standardise-key}
# prescription_id is the join key; lower-case and trim both sides.
ccmdd[, prescription_id := str_trim(str_to_lower(prescription_id))]
prescription_data[, prescription_id := str_trim(str_to_lower(prescription_id))]

# how many package prescription_ids have a matching prescription?
length(intersect(
  ccmdd$prescription_id,
  prescription_data$prescription_id
))
```

```{r merge}
setkey(ccmdd, prescription_id)
setkey(prescription_data, prescription_id)

# left join: keep every package and add prescription information where available.
packages_scripts_merged <- merge(
  ccmdd,
  prescription_data,
  by = "prescription_id",
  all.x = TRUE
)

# Step 2 contains one row per prescription_id, so the join should not
# increase the number of package rows.
stopifnot(nrow(packages_scripts_merged) == nrow(ccmdd))
```

```{r parse-dates}
# make sure all date columns are in the correct format.
setDT(packages_scripts_merged)

date_cols <- c(
  "date_created", "date_updated", "generate_time", "expected_delivery_time",
  "actual_delivery_time", "dispense_date", "return_date", "collection_date",
  "pup_scan_return_date", "pup_scan_support_location_date",
  "Registered date", "Prescription Date"
)

packages_scripts_merged[
  , (date_cols) := lapply(
    .SD,
    function(x) ymd(sub("\\s.*$", "", as.character(x)))
  ),
  .SDcols = date_cols
]
```

```{r summary}
skim(packages_scripts_merged)
```

# QC: prescriptions with no dates after the join

```{r qc-missing-dates}
# package rows where neither prescription date came through.
missing_dates_df <- packages_scripts_merged[
  is.na(`Registered date`) & is.na(`Prescription Date`)
]
nrow(missing_dates_df)

missing_ids <- packages_scripts_merged[
  is.na(`Registered date`) & is.na(`Prescription Date`),
  unique(prescription_id)
]
length(missing_ids)

# confirm these IDs are absent from the prescription set.
prescription_data[prescription_id %in% missing_ids]

# any of these IDs appearing on more than one package?
table(missing_dates_df$prescription_id)[
  table(missing_dates_df$prescription_id) > 1
]
```

# QC: distribution checks

```{r qc-distributions}
packages_scripts_merged %>%
  count(Gender, sort = TRUE) %>%
  mutate(percent = round(n / sum(n) * 100, 2))

packages_scripts_merged %>%
  count(`Prescription Length`, sort = TRUE) %>%
  mutate(percent = round(n / sum(n) * 100, 2))
```



# prescription_ids in packages but not in scripts

```{r qc-unmatched}
unmatched_ids <- setdiff(
  unique(ccmdd$prescription_id),
  unique(prescription_data$prescription_id)
)

length(unmatched_ids)
```

