---
title: "Step 7 — Adding coordinates and Facility↔Pickup-point distance"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(stringr)
library(tidyr)
library(readxl)
library(data.table)
library(skimr)
```

# Purpose

This file attaches coordinates to each package's `Facility` and `Pickup point`, then calculates the distance between the two as a proxy for how far a patient travels from their clinic to collect medication.

It uses the manually completed coordinate mapping produced after Step 6.

The `distance_euclid` measure is the straight-line Euclidean distance calculated directly from longitude and latitude degrees:

`sqrt((Δlatitude)^2 + (Δlongitude)^2)`

It is not a true geographic distance and is used only as a relative proxy for travel distance.

Inputs:
- `packages_scripts_merged` from Step 3
- `locations_match_audit._filled.xlsx` (Step 6 coordinate mapping after manual completion)

Output:
- `packages_with_distance` (package-level data with a computable Facility-to-Pickup-point distance)



# Configuration

```{r config}
# Step 3 must be run first to create the package-prescription object.
if (!exists("packages_scripts_merged")) {
  stop("Run Step 3 first to create `packages_scripts_merged`.")
}

# Manually completed coordinate mapping from Step 6.
# This file is included in the repository.
coords_xlsx <- "locations_match_audit._filled.xlsx"
```

# Load inputs

```{r load}
# coordinate mappings completed after Step 6.
facilities <- read_excel(
  coords_xlsx,
  sheet = "Facilities"
)

pickup_points <- read_excel(
  coords_xlsx,
  sheet = "PickupPoints"
)

# create a working copy of the package-level dataset from Step 3.
full <- copy(packages_scripts_merged)
setDT(full)
```

# Tidy the name fields

```{r tidy-names}
# treat "", "null", "n/a", "na" (any case) as missing.
nullish <- function(x) is.na(x) | tolower(trimws(x)) %in% c("", "null", "n/a", "na")

# drop packages with no patient_id (can't be used at person level downstream).
full <- full[!is.na(patient_id) & trimws(as.character(patient_id)) != ""]

# normalise placeholder blanks in the name fields to real NA.
full[, `:=`(
  Facility       = ifelse(nullish(Facility), NA_character_, Facility),
  `Pickup point` = ifelse(nullish(`Pickup point`), NA_character_, `Pickup point`)
)]
```

# Build coordinate lookups

```{r coord-lookups}
# coordinates may use a comma decimal separator; convert to a numeric.
to_num <- function(x) as.numeric(str_replace(as.character(x), ",", "."))

# one row per facility / pickup-point name (keep the first if duplicated).
fac_lu <- as.data.table(facilities)[
  , .(Facility = Name, fac_lon = to_num(Longitude), fac_lat = to_num(Latitude))
][, .SD[1], by = Facility]

pup_lu <- as.data.table(pickup_points)[
  , .(`Pickup point` = Name, pup_lon = to_num(Longitude), pup_lat = to_num(Latitude))
][, .SD[1], by = `Pickup point`]
```

# Join coordinates and compute distance

```{r distance}
# attach facility and pickup-point coordinates, then the eucl (degree-based)
# distance between them. Rows missing either coordinate get NA.
full_with_distance <- full %>%
  left_join(fac_lu, by = "Facility") %>%
  left_join(pup_lu, by = "Pickup point") %>%
  mutate(
    distance_euclid = ifelse(
      is.na(fac_lon) | is.na(fac_lat) | is.na(pup_lon) | is.na(pup_lat),
      NA_real_,
      sqrt((fac_lat - pup_lat)^2 + (fac_lon - pup_lon)^2)
    )
  ) %>%
  select(-fac_lon, -fac_lat, -pup_lon, -pup_lat)

setDT(full_with_distance)
message("Rows with no computable distance (missing a coordinate): ",
        sum(is.na(full_with_distance$distance_euclid)))
```

# Same site, no coordinates → distance 0

```{r same-site-zero}
# if the facility and pickup point have the same name but we couldn't compute a distance (a coordinate was missing), 
# the patient collects at their clinic, so distance is 0.
full_with_distance[Facility == `Pickup point` & is.na(distance_euclid),
                   distance_euclid := 0]

message("Rows with no computable distance after the same-site fill: ",
        sum(is.na(full_with_distance$distance_euclid)))
```

# Same site recorded under different names (QC + manual fix)

Some rows have a distance of approximately zero because the Facility and Pickup point have identical coordinates, but the two fields use different names for the same physical site.

These known variants are harmonised to a single canonical name.

```{r inspect-zero-distance}
# zero-distance rows whose two names differ: candidates for name harmonisation.
df_same_place_diff_name <- as.data.frame(full_with_distance[
  !is.na(distance_euclid) & abs(distance_euclid) < 1e-9 & Facility != `Pickup point`
])
nrow(df_same_place_diff_name)

# the distinct names involved (used to build the manual map below).
sort(unique(df_same_place_diff_name$Facility))
sort(unique(df_same_place_diff_name$`Pickup point`))
```

```{r manual-name-map}
# Manual name corrections for known variants of the same physical site.
# These were identified from the zero-distance pairs above.
# Each `from` name is replaced with the standard `to` name so that the same
# place is labelled consistently across the dataset.
manual_map <- data.table(
  from = c(
    "HD: Braemer Clinic",
    "HD : St Chads chc",
    "Eshowe Gateway", "Eshowe Hosp: Hospital Pharmacy", "Eshowe Hosp: Sinethemba Clinic Pharmacy",
    "EG Usher Gateway",
    "Wentworth Gateway",
    "Rietvlei Gateway",
    "Osindisweni Hosp: Main Pharmacy",
    "Appelsbosch Gateway", "Appelsbosch Mob 1", "Appelsbosch Mob 2",
    "Mseleni Gateway",
    "Umphumulo Gateway",
    "Verulam Medical Clinic _ Dr Lashika Athmaram",
    "St Andrew's Gateway",
    "St Apollinaris Mob 1", "St Apollinaris Mob 2",
    "Greytown Mob 1", "Greytown Mob 2",
    "Butterworth Gateway",
    "Sekororo Gateway",
    "Lebowakgomo Gateway",
    "Madadeni Gateway",
    "Mosvold Gateway",
    "HD : Nhlazatshe 4 Clinic",
    "C Johnson Mob 1", "C Johnson Mob 3",
    "Utrecht Mob 1", "Utrecht Mob 2", "Utrecht Mob 3",
    "Manguzi Mob 1", "Manguzi Mob 2", "Manguzi Mob 3",
    "St Margaret's TB MDR Hosp", "St Margaret's Clinic", "St Margaret's Gateway",
    "Sterkspruit Town Clinic", "Sterkspruit Crisis Centre",
    "Church Scot Mob 1", "Church Scot Mob 2"
  ),
  to = c(
    "Braemer Clinic",
    "St Chads CHC",
    "Eshowe Hosp", "Eshowe Hosp", "Eshowe Hosp",
    "EG Usher Mem Hosp",
    "Wentworth Hosp",
    "Rietvlei Hosp",
    "Osindisweni Hosp",
    "Appelsbosch Hosp", "Appelsbosch Hosp", "Appelsbosch Hosp",
    "Mseleni Hosp",
    "Umphumulo Hosp",
    "Verulam Clinic",
    "St Andrew's Hosp",
    "St Apollinaris", "St Apollinaris",
    "Greytown", "Greytown",
    "Butterworth Hosp",
    "Sekororo Hosp",
    "Lebowakgomo Hosp",
    "Madadeni Hosp",
    "Mosvold Hosp",
    "Nhlazatshe Clinic",
    "C Johnson", "C Johnson",
    "Utrecht", "Utrecht", "Utrecht",
    "Manguzi", "Manguzi", "Manguzi",
    "St Margaret's", "St Margaret's", "St Margaret's",
    "Sterkspruit", "Sterkspruit",
    "Church Scot", "Church Scot"
  )
)

# Replace a name with its canonical form (leave unmapped names unchanged).
recode_one <- function(x, map) map[.(x), on = .(from), ifelse(is.na(to), x, to)]

# Apply across the WHOLE dataset so every occurrence of a variant name is
# harmonised consistently (not only the zero-distance rows the map was built from).
full_with_distance[, `Pickup point` := recode_one(`Pickup point`, manual_map)]
full_with_distance[, Facility       := recode_one(Facility,       manual_map)]
```

# Check after name harmonisation

```{r check-after-map}
# after harmonising, check how many zero-distance rows there are
# and how many Facility and Pickup point names now match.
full_with_distance[, .(
  zero_distance = sum(
    !is.na(distance_euclid) &
      abs(distance_euclid) < 1e-9
  ),
  same_name = sum(
    !is.na(Facility) &
      !is.na(`Pickup point`) &
      Facility == `Pickup point`
  )
)]
```

# Create final Step 7 object

```{r final-object}
# keep only rows where a distance could be calculated or assigned.
packages_with_distance <- full_with_distance[
  !is.na(distance_euclid)
]

message(
  "Rows kept: ",
  nrow(packages_with_distance),
  " of ",
  nrow(full_with_distance),
  " (",
  nrow(full_with_distance) - nrow(packages_with_distance),
  " dropped)"
)
```

# Quick look

```{r summary}
skim(packages_with_distance)
```
